A North Sámi to South Sámi machine translation prototype

### 2014-02-09 paper/paper.tex
Added numbers from the oovtast presentations.
updates
update

\begin{itemize}
\item 3568 general word pairs + 61 proper nouns
\item added: 873 special domain word pairs 
\item added: 10 721 proper nouns
\end{itemize}

There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a general North Saami-Norwegian dictionary and a general South Saami-Norwegian dictionary. The word pairs were manually edited and the work revealed many incorrect pairs because of Norwegian words with more than one meaning. Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}. The result of this work was 3568 general word pairs and 61 proper nouns.

@@

The pivot model causes some extra lexical selection. Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}. But in South Saami there is no common verb for these concepts, and the lexical distribution is like the one for Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.

@@

WER -- word error rate
\begin{table}
  \begin{center}
    \begin{tabular}{|l|r|r|r|}
      \hline
      \textbf{ } & \textbf{Eval. A} & \textbf{Eval. B} & \textbf{Eval. C}  \\
      \hline
word order in account & 62,37  &  49,46  &  52,69 \\
word order not in account  &  36,20   &  28,32  &  28,32 \\
      \hline
    \end{tabular}
    \label{table:coverage}
    \caption{Vocabulary coverage of the system}
  \end{center}
\end{table}

@@

>>>>

\begin{itemize}
\item 3,568 general word pairs + 61 proper nouns
\item added: 873 special domain word pairs 
\item added: 10 721 proper nouns
\end{itemize}

There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a general North Saami-Norwegian dictionary and a general South Saami-Norwegian dictionary. The word pairs were manually edited and the work revealed many incorrect pairs because of Norwegian words with more than one meaning. Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}. The result of this work was 3,568 general word pairs and 61 proper nouns.

@@

The pivot model causes some extra lexical selection. Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calculate’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}. But in South Saami there is no common verb for these concepts, and the lexical distribution is like the one for Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.

@@

WER -- word error rate
\begin{table}
  \begin{center}
    \begin{tabular}{|l|r|r|r|}
      \hline
      & \textbf{Eval. A} & \textbf{Eval. B} & \textbf{Eval. C}  \\
      \hline
\textbf{WER} (\%) & 62.37  &  49.46  &  52.69 \\
\textbf{PER} (\%) &  36.20   &  28.32  &  28.32 \\
      \hline
    \end{tabular}
    \label{table:coverage}
    \caption{Evaluation metrics.}
  \end{center}
\end{table}

@@

From the oovtast-presentation:
\begin{table*} 
\begin{center}
\begin{tabular}{|l|r||r|r|r|}
\hline
\textbf{ }  & \textbf{MT} &  \textbf{A} & \textbf{B} & \textbf{C} \\
\hline
Lexical differences: & &   37 & 27 & 19 \\
\hline
Morphological differences: & &   15 & 14 & 16 \\
\hline
\hline
\multicolumn{5}{|l|} {\textbf{Syntax} }   \\
\hline
SOV/SAdvlV & 4 &  10 & 12 & 12 \\
\hline
Det pronomen & 4 &   7 & 4 & 7 \\
\hline
Ind. articles & 0 &   7 & 0 & 1 \\
\hline
Complement instead of genitive mod. & -- &   2 & 2 & 0 \\
\hline
Habitive \textit{utnedh} & 0 &  5 & 2 & 2 \\
\hline
\hline
All together & 8 &   31 & 20 & 21 \\
\hline
\end{tabular}
\end{center}
\end{table*}
